---
title: "North Carolina Public School Demographics (2014-2024)"
date: last-modified
date-format: "[Last modified on] MMMM DD, YYYY HH:mm:ss zzz"
format: 
  html:
    output: bookdown::html_document2
    fig_caption: yes
    theme: cosmo  # zephyr pulse, sketchy, default, cosmo, vapor etc
author:
  - name: Angus Victor Withee
    email: witheeav@appstate.edu
    affiliation:
      - name: Appalachian State University
        city: Boone
        state: NC  
        url: https://www.appstate.edu/
keywords:
  - R
  - data viz
license: "CC BY"  
code-fold: false
---
# Introduction

This is an exploration and visualization of North Carolina public school demographic data from February 2014 to February 2024. All data was obtained from the North Carolina Public School Statistical Profile (http://apps.schools.nc.gov/ords/f?p=145:1).

```{r}
#| label: "setup"
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = NA, fig.align = "center")
```

## Loading Necessary Packages

```{r}
#| label: "needed"
library(dplyr)
library(lubridate)
library(tidyr)
library(ggmap)
library(mapproj)
library(ggplot2)
library(caret)
```

## Importing Data
```{r}
demographics <- read.csv("~/STT2860/anguswithee.github.io/demographics.csv")
```

## Variables

`demographics`
```{r,echo=FALSE}
glimpse(demographics)
```

There are 17 variables in the `demographics` data frame.

1. `year`: The year the data was collected (2014-2024).
2. `LEA.Name`: The name of the county or counties in which the data was collected.
3. `Indian Male`: The number of Indigenous male students in the given county each year.
4. `Indian Female`: The number of Indigenous female students in the given county each year.
5. `Asian Male`: The number of Asian male students in the given county each year.
6. `Asian Female`: The number of Asian female students in the given county each year.
7. `Hispanic Male`: The number of Hispanic male students in the given county each year.
8. `Hispanic Female`: The number of Hispanic female students in the given county each year.
9. `Black Male`: The number of Black male students in the given county each year.
10. `Black Female`: The number of Black male students in the given county each year.
11. `White Male`: The number of White male students in the given county each year.
12. `White Female`: The number of White female students in the given county each year.
13. `Pacific Island Male`: The number of Pacific Island male students in the given county each year.
14. `Paicific Island Female`: The number of Hispanic female students in the given county each year.
15. `Two or More Male`: The number of male students who identify with two or more races in the given county each year.
16. `Two or More Female`: The number of female students who identify with two or more races in the given county each year.
17. `Total`: The total number of students in the given county each year.

## Data Wrangling

To make the data ideal for visualization, merge and rename variables.
```{r}
demographics <- demographics |> 
  group_by(Year,LEA.Name) |>
  summarize(aapi_male = ASIAN.Male+PACIFIC.ISLAND..Male,
            aapi_female = ASIAN.Female+PACIFIC.ISLAND..Female,
            indigenous_male = INDIAN.Male,
            indigenous_female = INDIAN.Female,
            hispanic_male = HISPANIC.Male,
            hispanic_female = HISPANIC.Female,
            black_male = BLACK.Male,
            black_female = BLACK.Female,
            white_male = WHITE.Male,
            white_female = WHITE.Female,
            multiracial_male = TWO.OR.MORE.Male,
            multiracial_female = TWO.OR.MOREFemale)
demographics <- demographics |> filter (Year>1)
```
Then, reshape the dataset from a wide format to a long format. 
```{r}
demographics <- demographics |> 
  select(Year, LEA.Name, 
         aapi_male, aapi_female, 
         indigenous_male, indigenous_female, 
         hispanic_male, hispanic_female, 
         black_male, black_female, 
         white_male, white_female, 
         multiracial_male, multiracial_female) |> 
  pivot_longer(
    cols = starts_with("aapi") | starts_with("indigenous") | 
           starts_with("hispanic") | starts_with("black") | 
           starts_with("white") | starts_with("multiracial"),
    names_to = c("race", "sex"),
    names_pattern = "(.*)_(male|female)",
    values_to = "count"
  )
demographics$Year <- as.numeric(demographics$Year)
head(demographics)
```
# Data Visualization

## NC Public School Demographics (2024)

Before we look at changes over time, let's start by looking at demographics in 2024.

```{r}
demo2024 <- demographics |>
  filter(Year==2024)
```

```{r fig1, fig.cap="Figure 1: NC Public School Students by Race (2024)."}
ggplot(demo2024, aes(x = race, y = count, fill = race)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Race",
    y = "Number of Students"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) + guides(fill="none") +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1))
```

NC Public School Demographics by Sex (2024)

```{r fig2, fig.cap="Figure 2: NC Public School Demographics by Sex (2024)."}
ggplot(demo2024, aes(x = sex, y = count, fill = sex)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Sex",
    y = "Number of Students"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1))
```

```{r fig3, fig.cap="Figure 3: NC Public School Demographics by Race and Sex (2024)."}
ggplot(demo2024, aes(x = race, y = count, fill = sex)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Race",
    y = "Number of Students"
  ) + 
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1))
```

## NC Public School Demographics Over Time

```{r}
demographic_trends <- demographics |> 
  group_by(Year, race, sex) |> 
  summarize(total_count = sum(count, na.rm = TRUE), .groups = "drop")
```

```{r fig4, fig.cap="Figure 4: NC Public School Racial Demographic Trends (2024)."}
ggplot(demographic_trends, aes(x = Year, y = total_count, color = race)) +
  geom_smooth(method = "loess", se = FALSE, size = 1)+
  labs(
    x = "Year",
    y = "Total Count"
  ) +
  scale_x_continuous(breaks = unique(demographic_trends$Year)) + # Show all years
  scale_y_continuous(labels = scales::comma) +
    theme_minimal() +
    theme(panel.border = element_rect(color = "black", fill = NA, size = 1))
```

```{r fig5, fig.cap="Figure 5: NC Public School Race and Sex Demographic Trends (2014-2024)."}
ggplot(demographic_trends, aes(x = Year, y = total_count, color = race, linetype = sex)) +
  geom_line(size = 1) +
  labs(
    x = "Year",
    y = "Total Count"
  ) +
  scale_x_continuous(breaks = unique(demographic_trends$Year)) + # Show all years
  scale_y_continuous(labels = scales::comma) +
    theme_minimal() +
    theme(panel.border = element_rect(color = "black", fill = NA, size = 1))
```

```{r}
white_total <- demographic_trends |> 
  filter(race == "white") |> 
  group_by(Year) |> 
  summarize(white_count = sum(total_count), .groups = "drop")

total_count_by_year <- demographic_trends |> 
  group_by(Year) |> 
  summarize(total_students = sum(total_count, na.rm = TRUE), .groups = "drop")

poc_data <- left_join(total_count_by_year, white_total, by = "Year") |> 
  mutate(poc_percentage = 100 * (total_students - white_count) / total_students)
```

```{r fig6, fig.cap="Figure 6: NC Public School Percentage of Students of Color Over Time (2014-2024)."}
ggplot(poc_data, aes(x = Year, y = poc_percentage)) +
  geom_line(size = 1) +
  labs(
    x = "Year",
    y = "Percentage of Students of Color"
  ) +
  theme_minimal() +
  scale_y_continuous(limits = c(0, 100),labels = scales::percent_format(scale = 1)) +
    scale_x_continuous(breaks = unique(demographic_trends$Year))
```

## Watauga County Public School Demographics

```{r}
wata_demo <- demographics |>
  filter(LEA.Name=="Watauga County Schools")

wata_demo2024 <- wata_demo |>
  filter(Year==2024)

wata_demo_trends <- wata_demo |>
  group_by(Year, race, sex) |> 
  summarize(total_count = sum(count, na.rm = TRUE), .groups = "drop")

wata_white_total <- wata_demo_trends |>
  filter(race == "white") |> 
  group_by(Year) |> 
  summarize(white_count = sum(total_count), .groups = "drop")

wata_total_count_by_year <- wata_demo_trends |>
  group_by(Year) |> 
  summarize(total_students = sum(total_count, na.rm = TRUE), .groups = "drop")

wata_poc_data <- left_join(wata_total_count_by_year, wata_white_total, by = "Year") |> 
  mutate(poc_percentage = 100 * (total_students - white_count) / total_students)
```

```{r fig7, fig.cap="Figure 7: Watauga County Public School Demographics by Race and Sex (2024)."}
ggplot(wata_demo2024, aes(x = race, y = count, fill = race)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Race",
    y = "Number of Students"
  ) +
  theme_minimal() + 
  guides(fill="none") +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1))
```

```{r fig8, fig.cap="Figure 8: Watauga County Public School Race and Sex Demographics Over Time (2014-2024)"}
ggplot(wata_demo_trends, aes(x = Year, y = total_count, color = race, linetype = sex)) +
  geom_line(size = 1) +
  labs(
    x = "Year",
    y = "Total Count"
  ) +
  scale_x_continuous(breaks = unique(demographic_trends$Year)) + # Show all years
  scale_y_continuous(labels = scales::comma) +
    theme_minimal() +
    theme(panel.border = element_rect(color = "black", fill = NA, size = 1))
```

```{r fig9, fig.cap="Figure 9: Watauga County Public School Percentage of Students of Color Over Time (2014-2024)."}
ggplot(wata_poc_data, aes(x = Year, y = poc_percentage)) +
  geom_line(size = 1) +
  labs(
    x = "Year",
    y = "Percentage of Students of Color"
  ) +
  theme_minimal() +
  scale_y_continuous(limits = c(0, 100),labels = scales::percent_format(scale = 1)) +
    scale_x_continuous(breaks = unique(demographic_trends$Year))
```
